---
title: "IntroR"
author: "Liliana Lopez-Kleine"
date: "28 de noviembre de 2017"
output:
  pdf_document: default
  html_document: default
---

This is a brief introduction on basic data manipulation and on how to perform basic statistical analysis using R.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

R is an open source program and has many contributors
```{r }
licence()
contributors()
```


This command stands for get working directory and gives you the directory in which R is currently working (where it will search for the files you upload or where it will save everything)
```{r }
getwd()
```
For this tutorial to work you should change it to the folder Rarchivos. You can use the command or do it via clicks (in Rstudio it will be Session - Set working directory - choose directory)
```{r }
setwd("~/BioestadisticaGeneral/R/Rarchivos")
```


If you do not want the code to be printed out in your output document add echo=FALSE as for example:

```{r echo=FALSE}
getwd()
```

Now we will start creating simple objects in R:
```{r }
w<-1:25
z <- scan()#Hit enter two times to stop 
x <- c(3,8,9,6,4,5) 
w <- 1:10 
w%in%x
ind=which(w%in%x)
wx<-w[ind]
y <- seq(1, 20, 2) 
u<-rep(1,7)
u<-rep(c(1,2),c(3,4))
t<- rep(c(3,4),c(3,4))
ut1<-cbind(u,t) #combine columns
ut2<-rbind(u,t) #combine rows
class(ut2)
x <- rnorm(1000, mean = 3, sd = 2) 
x <- matrix(x, nrow = 10, ncol = 100) 
class(x)
x2<-as.data.frame(x)
class(x2)
y <- x[3, ] 
y <- x[, -1] 
y <- x[x >= 4] 

```

If you know the function name and want to get access to the help page

```{r }
#?matrix
```
If you do not know the name of the function use 
```{r }
help.search("matrix")
```
You can also create a data frame loading up data from other formats as for example txt.

```{r }
setwd("~/BioestadisticaGeneral/R/Rarchivos")
```
```{r }
read.table("t3var.txt", header=TRUE) 
```

The table is uploaded but we have not created an object that contains it, to do so and make some descriptions of the variables
```{r }
t3var<-read.table("t3var.txt", header=TRUE) 
summary(t3var)
```
```{r }
plot(t3var$peso, t3var$talla)
```
If you want to extract some data you can use boolean operators like "=="
```{r }
x <- t3var$talla[t3var$sexo == "h"]
y <- t3var$peso[t3var$sexo == "h"]
plot(x, y)
hist(x)
hist(y)
```
```{r }
x <- t3var$talla[t3var$sexo == "h"]
y <- t3var$peso[t3var$sexo == "h"]
plot(x, y)#plot of men's data
```
```{r }
par(mfrow = c(2, 2))#For plotting more than one plot on the same window
hist(x)
hist(y)
qqnorm(x)# a graphical way of testing adjustment to the normal distribution
qqnorm(y)

```

Suppose you want to save the men's data. Create a matrix and save it as it is or as a data frame.
```{r }
M<-cbind(x,y)
dput(M, "Mensdata.csv")#use deget("Mensdata.csv") to load again
write.table(M, "MensdataDF.txt")#Use read.table to load again
```
R will recognize numerical data vectors and categorical data vectors as for example sexo of t2var

```{r }
class(t3var$sexo)
attach(t3var)#this command allows R to recognize the objects in the data frame
class(peso)
```
If a variable is coded with numbers, there can be a confusion
```{r }
sitio<-c(1,1,1,1,1,1,1,1,1,0,0,0,0,0,0,0,0,0)
summary(sitio)
summary(as.factor(sitio))
sitio<-c(1,1,1,1,1,1,1,1,1,0,0,0,0,0,0,0,0,0,NA,NA,NA)
summary(as.factor(sitio))
summary(as.factor(na.omit(sitio)))
pie(summary(as.factor(na.omit(sitio))))
barplot(summary(as.factor(na.omit(sitio))))
barplot(summary(as.factor(na.omit(sitio))), horiz = TRUE, col="blue")
```

Outlier detection
```{r }
boxplot(talla) 
boxplot.stats(talla)
ind<-talla==200
t3var[ind,]
ind<-talla>=189
length(ind)
t3var[ind,]

#Changing the interquartile range to detect extreme outliers
boxplot(talla,range=3)
plot(lm(talla ~ 1))

```

```{r }
#Multivariate outlier detection (here bivariate)
clmedians<-c(median(t3var$peso),median(t3var$talla))
mahalanobis(t3var[,2:3],mean(clmedians),cov(t3var[,2:3]))->mahat3var
#centers are the medians
length(mahat3var) 
dim(t3var) 
boxplot(mahat3var)
boxplot.stats(mahat3var)

```


Now we are going to see how to handle a big data table, a gene expression table, for which an auxiliary annotation variable is available. The aim is to extract genes that are known to be resistance genes.
```{r }
tom2<-read.table("base-filtro3-6sept.txt",h=T) #tomato gene expression data
names(tom2) #in genes.filter the ID of the genes is found
attach(tom2)
genesR<-read.table("genes-resist.txt",h=T) #table indicating which genes are resistance genes (1) and which are not (0). 
dim(genesR)
names(genesR) #in genes.ori is the ID of all genes
attach(genesR)
genes.resistance<-genes.ori[resistance==1] #extracts only names of resistance genes
length(genes.resistance)
indres<-which(genes.filter%in%genes.resistance) # finds these genes in the original gene expression table
genesres<-genes.filter[indres] #names of the resistance genes
length(genesres)
tomresist<-tom2[genes.filter[indres],c(15:22)] #extract data of the resistance genes. Only the variables of interest (log2 transformed data)
dim(tomresist)
write.table(tomresist,"tomresist.csv")#saves the new table
```
Some descriptive statistics
```{r }
x<-c(1,2,3,4,5,6,6,7,8,10)
mean(x)
var(x)
sqrt(var(x))
sd(x)
median(1:4)
range(x)
quantile(x)
```

Correlation coefficient
```{r }
cor(tomresist)#correlation between samples
cor(tomresist,method="spearman")
cor.test(tomresist[,1],tomresist[,2])#significance of correlation
```

A plot of the correlation in library corrplot
```{r }
#install.packages("corrplot")
library(corrplot)
corM<-cor(tomresist)
corrplot(corM,method = "circle")#all correlations are very high because they are replicates
corrplot(corM,method = "square")
```

Hypothesis testing

T-test
```{r }
#?t.test
t.test(1:10,mu=5) 
t.test(1:10,y<-c(7:20)) 
t.test(1:10,y<-c(7:20), var.equal=TRUE) # eliminates the Welch correction on degrees of freedom 
t.test(1:10,y<-c(7:20, 200))#T test does not work because normality assumption is not met
wilcox.test(1:10,y<-c(7:20, 200))
```

Chi-square test for independence
```{r }
sexe <- as.factor(rep(c("F", "M"), c(10, 10)))
examen <- as.factor(c("R", "R", "R", "R", "E", "E", "E","E", "E", "E", "R", "R", "R", "R", "R", "R", "E", "E", "E", "E"))
table(sexe,examen)
chisq.test(sexe, examen)
chisq.test(sexe, examen, correct = FALSE)#only if more than one degree of freedom
```
Fisher exact test when frequencies of the contingency table are less than 3
```{r }
TeaTasting <-
matrix(c(3, 1, 1, 3),
       nrow = 2,
       dimnames = list(Guess = c("Milk", "Tea"),
                       Truth = c("Milk", "Tea")))
fisher.test(TeaTasting, alternative = "greater")
```

Estimate power of a T test or fix a power and estimate sampling size. d is the effect size as described by Cohen.
```{r }
library(pwr)

pwr.t.test(d=0.2,n=60,sig.level=0.10,type="one.sample",alternative="two.sided")

pwr.t.test(d=0.2,power=0.90,sig.level=0.10,type="one.sample",alternative="two.sided")

pwr.t.test(d=0.2,power=0.80,sig.level=0.010,type="one.sample",alternative="two.sided")
```

Power for analysis of variance
```{r }
power.anova.test(groups = 3, n = NULL, between.var = 43715, within.var = 102030, sig.level = 0.05,
power = 0.8)
```

Linear Regression
```{r }
MicroYprot<-read.table("MicroYProtprom.txt",h=T)
summary(MicroYprot)#Are there missing values?
attach(MicroYprot)
plot(promProt,promMicro)
```

```{r }
lmp<-lm(promProt~promMicro)#missing values are not used
summary(lmp)
#The quality of the model is very bad
protpred<-2.2936*promMicro[205]-1.0195#Predicting a missing value
protpred
```
```{r }
#Testing for assumptions
#normal distribution of residuals
plot(lmp) #several plots are displayed
shapiro.test(lmp$residuals)
```

```{r }
#independence
#install.packages("lmtest")
library(lmtest)
dwtest(promProt[1:204]~promMicro[1:204])
```

Multiple linear regression
```{r }
RLM<-read.table("RegrLinealMultiple.txt",h=T)
RLM
X<-RLM[1:13,3:5]
X<-as.matrix(X)
y<-RLM[1:13,1]
t(X)


lmR<-lm(y~X)
lmR
summary(lmR)#this is a good model
```

Analysis of variance
```{r }
dosis<-read.table("dosis.txt",h=T)
attach(dosis)

bartlett.test(Vigilancia~Dosis)

by(Vigilancia, Dosis, shapiro.test)


ks.test(Vigilancia[1:6], "pnorm",mean(Vigilancia[1:6]),sd(Vigilancia[1:6]))#Kolmogorov-Smirnov test for normality
```

```{r }
#1 Factor ANOVA
aov.dosis<- aov(Vigilancia~Dosis,data=dosis)
summary(aov.dosis)

#Multiple testing
pairwise.t.test(Vigilancia,Dosis)
pairwise.t.test(Vigilancia,Dosis,p.adjust="bonf")
```

```{r }
#2 Factor ANOVA
aov.dosis<- aov(Vigilancia~Genero+Dosis,data=dosis)
summary(aov.dosis)
aov.dosis<- aov(Vigilancia~Genero+Dosis+Genero*Dosis,data=dosis)
summary(aov.dosis)
```
```{r }
#Another way of multiple testing 
TukeyHSD(aov.dosis, "Dosis", ordered = TRUE)

plot(TukeyHSD(aov.dosis, "Dosis"))
```

If the normality assumption is not met, a 1 Factor ANOVA can be replaced by the non-parametric Kruskal-Wallis test
```{r }
kruskal.test(Vigilancia~Dosis,data=dosis)
```

For the experimental design, in order to assure randomization, the following library can be used
```{r }
#install.packages("agricolae")
library(agricolae)
trt<-c(3,2) # factorial 3x2 
outdesign <-design.ab(trt, r=3, serie=2) 

# factorial 2 x 2 x 2 with 5 replications in completely randomized design. 
trt<-c(2,2,2) 
outdesign<-design.ab(trt, r=5, serie=2,design="crd") #complete randomized design

# factorial 3 x 3 in latin square design. 
trt <-c(3,3) 
outdesign<-design.ab(trt, serie=2, design="lsd") #latin square design

#incomplete blocks
# 4 treatments and k=3 size block 
trt<-c("A","B","C","D") 
k<-3 
outdesign<-design.bib(trt,k,serie=2,seed =41) # seed = 41 print(outdesign$parameters) 
#Marsaglia's famous Super-Duper method to generate random numbers as default
book<-outdesign$book 
plots <-as.numeric(book[,1]) 
matrix(plots,byrow=TRUE,ncol=k) 
print(outdesign$sketch)
```

Some useful functions

```{r }
#apply
	x <- cbind(x1 = 3, x2 = c(4:1, 2:5))
	dimnames(x)[[1]] <- letters[1:8]
	#?apply
	col.sums <- apply(x, 2, sum)
	col.sums <- apply(x, 1, sum)
	lapply(x, mean)
 sapply(x, mean)
```

Constructing loops
```{r }
x12<- c(5,4,3,2)
	for(i in 1:3){x12[i] <- x12[i] + 1}
	x12
	#[1] 6 5 4 2
	x12 <- x12 + 1
	for(i in 1:length(x12)){x12[i] <- x12[i] + 1}
	for(i in 1:length(x12)){if(x12[i] < 4){x12[i] <- x12[i] + 1}}

  	if ( x > 3) { resultado <- "x es mayor de 3"}   # Efectúa la comparación.
```

```{r }
#A simple function
	x<-5
	f <- function(x) {
 	 x^2 + x + 1
	}
	f(x)
```

